\LaTeXe\ guide for the Journal of Functional Programming

### 2019-04-27 papers/jfp/8-engineering.tex
Finish S8.7: Self-tracking

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
data Key k v s a where
    Script :: k -> Key k v s s -- Keys for build scripts
    Value  :: k -> Key k v s v -- Keys for all other values

selfTracking :: (s -> Task Monad k v) -> Tasks Monad k s
                                      -> TasksT Monad (Key k v s)
selfTracking _     _     (Script _) = Nothing -- Scripts are inputs
selfTracking parse tasks (Value  k) = runScript <$> tasks k
  where
    -- Fetch the script, parse it, and then run the obtained task
    runScript :: Task Monad k s -> TaskT Monad (Key k v s) v
    runScript task = TaskT $ \fetch -> do
        script <- run task (fetch . Script)
        run (parse script) (fetch . Value)
\end{minted}
\vspace{1mm}

>>>>

Below we show how to implement self-tracking in a build system that allows users
to describe build tasks by \emph{scripts} written in a non-embedded domain
specific language. We will denote the type of scripts by \hs{s}, and will assume
that scripts are indexed by keys \hs{k} just like all other values \hs{v}. More
specifically, we use the following GADT to tag keys \hs{k} with corresponding
result types: \hs{s} for scripts, and \hs{v} for other values.

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
data Key k v s a where
    Script :: k -> Key k v s s -- Keys for build scripts
    Value  :: k -> Key k v s v -- Keys for all other values
\end{minted}
\vspace{1mm}

\noindent
The function \hs{selfTracking} defined below is a generalisation of the approach
explained in the above \Excel example \hs{sprsh4}. The function takes a parser
for scripts, of type \hs{s}~\hs{->}~\hs{Task}~\hs{Monad}~\hs{k}~\hs{v}, and a
description of \emph{how to build all scripts}, of type
\hs{Tasks}~\hs{Monad}~\hs{k}~\hs{s}. For \hs{sprsh4}, the latter would simply
fetch \cmd{B1-formula} when given the key \cmd{B1} and return \hs{Nothing}
otherwise, but the presented approach can cope with much more sophisticated
scenarios where scripts themselves are derived from ``script sources'', e.g. all
C compilation scripts can be obtained from a single pattern rule, such as
\cmd{gcc -c [source] -o [object]}. The resulting typed task description
\hs{TasksT}~\hs{Monad}~\hs{(Key}~\hs{k}~\hs{v}~\hs{s)} tracks both values and
scripts that compute them, and is therefore self-tracking.

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
selfTracking :: (s -> Task Monad k v) -> Tasks Monad k s
                                      -> TasksT Monad (Key k v s)
selfTracking _     _     (Script _) = Nothing -- Scripts are inputs
selfTracking parse tasks (Value  k) = runScript <$> tasks k
  where
    -- Fetch the script, parse it, and then run the obtained task
    runScript :: Task Monad k s -> TaskT Monad (Key k v s) v
    runScript task = TaskT $ \fetch -> do
        script <- run task (fetch . Script)
        run (parse script) (fetch . Value)
\end{minted}
\vspace{1mm}

\noindent
It is possible to implement \hs{selfTracking} without relying on GADTs and typed
tasks presented in~\S\ref{sec-polymorphism}, but the cost of using partial
functions.
